---
title: "MOFA MCMC"
author: "Frederik Ziebell"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache= TRUE, collapse = TRUE)
```

# Preparations
```{r}
suppressPackageStartupMessages({
  library("MOFA2")
  library("greta")
  library("cowplot")
  library("tidyverse")
})
source("helper_functions.R")
theme_set(theme_cowplot())
```

# Load data
We load the MOFA model tained in `1_mofa_original.Rmd` and extract the $W$ and $Z$ matrices.
```{r}
mofa_obj <- load_model("mofa_obj_trained.hdf5")
Y_drug <- mofa_obj@data$drug$group1
Y_met <- mofa_obj@data$met$group1
Y_rna <- mofa_obj@data$rna$group1

Z_mofa <- mofa_obj@expectations$Z$group1
W_drug_mofa <- mofa_obj@expectations$W$drug
W_met_mofa <- mofa_obj@expectations$W$met
W_rna_mofa <- mofa_obj@expectations$W$rna
```

As observed in the full dataset, there is good overlap between the different views regarding explained variance. Factor 1 separates patients by IGHV status. Apart from factors 1 and 2, there is little correlation between the factors.
```{r}
plot_variance_explained(mofa_obj)
plot_factor_cor(mofa_obj)
plot_factors(mofa_obj, color_by="IGHV")
```

We check the approximation quality of the matrix factorizations by computing the observed vs. predicted residuals of the data matrices.
```{r, fig.width=10}
W_drug_mofa <- mofa_obj@expectations$W$drug
W_met_mofa <- mofa_obj@expectations$W$met
W_rna_mofa <- mofa_obj@expectations$W$rna
Z_mofa <- mofa_obj@expectations$Z$group1

plot_grid(
  qplot(Y_drug - tcrossprod(W_drug_mofa,Z_mofa), bins=100) + labs(x="residuals", title="drug"),
  qplot(Y_met - tcrossprod(W_met_mofa,Z_mofa), bins=100) + labs(x="residuals", title="met"),
  qplot(Y_rna - tcrossprod(W_rna_mofa,Z_mofa), bins=100) + labs(x="residuals", title="rna"),
  nrow = 1
)
```

# Define model
We implement the MOFA model with an ARD prior on the weights matrices, but no spike-and-slab prior. This is consistent with the loaded MOFA model, as stated in the `mofa_obj@model_options` slot.
```{r}
# =====================================
# dimensions
# =====================================
# number of samples
N <- ncol(Y_drug) 
# number of factors
K <- ncol(Z_mofa) 
# number of features
D_drug <- nrow(Y_drug)
D_met <- nrow(Y_met)
D_rna <- nrow(Y_rna)

# =====================================
# data
# =====================================
# greta assumes observations in rows and variables in columns
Yt_drug <- t(Y_drug)
Yt_met <- t(Y_met)
Yt_rna <- t(Y_rna)

# =====================================
# hyperparameters
# =====================================
# Note: mofapy2 uses the shape = a, rate = b parametrization of the 
# Gamma distribution. This corresponds to the alpha = a, beta = b 
# parametrization of extraDistr::dinvgamma used by greta.
a0_alpha <- 1e-3
b0_alpha <- 1e-3
a0_tau <- 1e-14
b0_tau <- 1e-14

# ====================================
# priors
# ====================================

# creates a D x K greta array that defines the prior of a W matrix
define_W_prior <- function(D, K) {
  one_over_alpha <- inverse_gamma(alpha = rep(a0_alpha, K), beta = rep(b0_alpha, K))
  # D x K matrix where each row is a copy of one_over_alpha
  one_over_alpha_mat <- t(greta_array(rep(one_over_alpha, D), dim = c(K, D)))
  # W_hat
  normal(mean = zeros(D, K), sd = one_over_alpha_mat)
}

# creates a D x D greta array that defines the residual SD prior
# of a data matrix Y with D features
define_residual_prior <- function(D) {
  Sigma_Y <- zeros(D, D)
  diag(Sigma_Y) <- inverse_gamma(a0_tau, b0_tau, dim = D)
  Sigma_Y
}

# Z and W
Z <- normal(0, 1, dim = c(N, K))
W_drug <- define_W_prior(D_drug, K)
W_met <- define_W_prior(D_met, K)
W_rna <- define_W_prior(D_rna, K)

# residuals
Sigma_Y_drug <- define_residual_prior(D_drug)
Sigma_Y_met <- define_residual_prior(D_met)
Sigma_Y_rna <- define_residual_prior(D_rna)

# =====================================
# likelihood
# =====================================
distribution(Yt_drug) <- multivariate_normal(Z %*% t(W_drug), Sigma_Y_drug)
distribution(Yt_met) <- multivariate_normal(Z %*% t(W_met), Sigma_Y_met)
distribution(Yt_rna) <- multivariate_normal(Z %*% t(W_rna), Sigma_Y_rna)

# =====================================
# model
# =====================================
m <- model(Z, W_drug, W_met, W_rna)
```

# MCMC sampling
```{r}
set.seed(1)
draws <- mcmc(m, one_by_one = T, verbose = FALSE)
```

# Chain convergence
We assess the convergence of the four chains by randomly selecting 10 elements per matrix and inspecting their traceplots. There is almost no mixing between the chains, indicating that the parameters of the model are not identifiable, leading to chains that attain different local optima of the posterior.
```{r, fig.width=15, fig.height=15}
set.seed(5)
c(
  str_c("W_drug[",sample(1:nrow(W_drug),10,replace=T),",",sample(1:ncol(W_drug),10,replace=T),"]"),
  str_c("W_met[",sample(1:nrow(W_met),10,replace=T),",",sample(1:ncol(W_met),10,replace=T),"]"),
  str_c("W_rna[",sample(1:nrow(W_rna),10,replace=T),",",sample(1:ncol(W_rna),10,replace=T),"]"),
  str_c("Z[",sample(1:nrow(Z),10,replace=T),",",sample(1:ncol(Z),10,replace=T),"]")
) %>%
  bayesplot::mcmc_trace(draws, pars=.,facet_args = list(ncol = 5))
```


# Explore MCMC model
The factors of the MCMC model fail to explain variance and to disentangle the main source of variation in the data related to IGHV status. This is potentially due to the *rotational invariance* problem mentioned in the MOFA papers in conjunction with poor initial values. 
```{r}
mofa_obj_mcmc <- update_Z_and_W(mofa_obj, draws)
plot_variance_explained(mofa_obj_mcmc, use_cache = FALSE)
plot_factor_cor(mofa_obj_mcmc)
plot_factors(mofa_obj_mcmc, color_by="IGHV")
```

Moreover, the MCMC model gives a worse solution of the factorization problem, as the distribution of the residuals is much wider than the ones obtained by MOFA. 
```{r, fig.width=10}
W_drug_mcmc <- mofa_obj_mcmc@expectations$W$drug
W_met_mcmc <- mofa_obj_mcmc@expectations$W$met
W_rna_mcmc <- mofa_obj_mcmc@expectations$W$rna
Z_mcmc <- mofa_obj_mcmc@expectations$Z$group1

plot_grid(
  qplot(Y_drug - tcrossprod(W_drug_mcmc,Z_mcmc), bins=100) + labs(x="residuals", title="drug"),
  qplot(Y_met - tcrossprod(W_met_mcmc,Z_mcmc), bins=100) + labs(x="residuals", title="met"),
  qplot(Y_rna - tcrossprod(W_rna_mcmc,Z_mcmc), bins=100) + labs(x="residuals", title="rna"),
  nrow = 1
)
```


# MCMC with MOFA initial values
To assess the general validity of the MCMC implementation, we run the model again, but this time with the data matrices obtained from MOFA as initial values. The mixing between the chains is greatly improved.
```{r, fig.width=15, fig.height=15}
set.seed(1)
inits <- initials(Z = Z_mofa, W_drug = W_drug_mofa, W_met = W_met_mofa, W_rna = W_rna_mofa)
draws <- mcmc(m, one_by_one = T, initial_values = inits, verbose = FALSE)

set.seed(5)
c(
  str_c("W_drug[",sample(1:nrow(W_drug),10,replace=T),",",sample(1:ncol(W_drug),10,replace=T),"]"),
  str_c("W_met[",sample(1:nrow(W_met),10,replace=T),",",sample(1:ncol(W_met),10,replace=T),"]"),
  str_c("W_rna[",sample(1:nrow(W_rna),10,replace=T),",",sample(1:ncol(W_rna),10,replace=T),"]"),
  str_c("Z[",sample(1:nrow(Z),10,replace=T),",",sample(1:ncol(Z),10,replace=T),"]")
) %>%
  bayesplot::mcmc_trace(draws, pars=.,facet_args = list(ncol = 5))
```

Moreover, the second run of the model results in a factorization very close to the original one found by MOFA, indicating that the solution of MOFA is at a location of the posterior that is very hard to escape by the MCMC sampler.
```{r}
mofa_obj_mcmc <- update_Z_and_W(mofa_obj, draws)
plot_variance_explained(mofa_obj_mcmc, use_cache = FALSE)
plot_factor_cor(mofa_obj_mcmc)
plot_factors(mofa_obj_mcmc, color_by="IGHV")
```


# Posterior optimization
Finally, we use greta's capability to obtain the MAP via optimization. The MAP model explains more variance than the MCMC model without initial values. Factors are largely uncorrelated and there is some separation of patients by IGHV status.
```{r}
set.seed(1)
opt_res <- opt(m, optimiser = adam())

mofa_obj_opt <- update_Z_and_W_opt(mofa_obj, opt_res)
plot_variance_explained(mofa_obj_opt, use_cache = FALSE)
plot_factor_cor(mofa_obj_opt)
plot_factors(mofa_obj_opt, color_by="IGHV")
```